---
title: "Final Project"
author: "Caitlin Buckle"
date: "2026-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

#### Load required packages 
```{r message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(splines)
library(broom)
``` 


#### Datasets 

Shark Incident Data downloaded from Taronga Conservation Society, Public Access Australian Shark Incident Database (ASID)

Accessed and downloaded, July 16th 2026 


Population Data downloaded from Australian Bureau of Statistics (ABS) 

1901 - 2011: ASB 3105.0.65.001 Australian Historical Population Statistics, 2014 Table 1.2 Population by sex, states, and territories, 1901 onwards

2012 - 2019: ASB 3101.0 National State and Territory Population, Table 4: Estimated Resident Population, States and Territories 

2020 - 2025: ABS 3101.001.202512 National, State, and Territory Population, Dec 2025  Table 5: Population by sex-States and Territories

Accessed and downloaded, July 17th 2026 

### Read Data into R

```{r, results='hide'}
asid<-read.csv("Data/ASID.csv") #Australian Shark Incident Database file 
summary(asid)

pop_1901_2011<- read_csv("Data/1901_2011.csv") #Population 1901-2011, States and Territories 
pop_2012_2019 <- read_csv("Data/2012_2019.csv") #Population 2012-2019, States and Territories 
pop_2020_2025 <- read_csv("Data/2020_2025.csv") #Population 2020-2025, States and Territories 

```

### Cleaning - ASID 

```{r,results='hide'}
# STATE COLUMN 
asid%>% 
  count(State, sort = TRUE) #7 states/territories in data set, three entries with no state/territory (need to remove), and one Qld (instead of QLD - will amend)

asid$State[asid$State == "Qld"] <- "QLD" #changes Qld into QLD

asid<-asid%>%
  filter(!is.na(State) & State != "" & State != " ") #filters State column for any NA value, "" , or " " and removes

asid%>% 
  count(State, sort = TRUE) #State data is now all clean, with 7 states/territories with incidents in the database 

# VICTIM INJURY COLUMN 
asid%>% 
  count(Victim.injury, sort = TRUE) #'injured', 'fatal', and 'uninjured' the key categories, three for "blank" and two 'unknown' (will remove), one is 'injured' and one 'injury' (will include into 'injured')

asid$Victim.injury[asid$Victim.injury == "Injured"] <- "injured"
asid$Victim.injury[asid$Victim.injury == "injury"] <- "injured"

asid<-asid%>%
  filter(!is.na(Victim.injury) & Victim.injury != "" & Victim.injury != " " & Victim.injury != "unknown") #remove blank values and unknowns 

asid%>%
  count(Victim.injury,sort = TRUE) #now three categories - injured, fatal, or uninjured 

# INCIDENT YEAR COLUMN 
summary(asid$Incident.year) #dataset starts in 1791 and latest data value entry was in 2026 

sum(is.na(asid$Incident.year)) #no incidents are missing year - great!! 
```

### Cleaning - Population Statistics 

```{r,results='hide'}
# population data was all in different formats (some vertical orientation, some horizontal)
# use pivot to alter the orientation and then bind the different data frames into one population data frame 

clean_1901_2011 <- pop_1901_2011 %>%  
  rename(State = Year) %>% # need to rename the Year column to State (incorrect heading from original .csv file)
    pivot_longer(   # pivot all the year columns down, leaving State alone
    cols = -State,               
    names_to = "Year", 
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(Year))

clean_2012_2019 <- pop_2012_2019 %>%
  pivot_longer(
    cols = -Year,                
    names_to = "State", 
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(Year))

clean_2020_2025 <- pop_2020_2025 %>%
  pivot_longer(
    cols = -Year,                
    names_to = "State", 
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(Year))

final_population_df <- bind_rows( # join together the rows in the separate files into one data frame with all population data 1901-2025
    clean_1901_2011, 
    clean_2012_2019, 
    clean_2020_2025
  ) %>%
  mutate(State = toupper(State)) %>% #ensure column headers match the ASID headers to avoid confusion later on 
  arrange(State, Year)

summary(final_population_df)
```

### Summary: Total Incidents by State

```{r,results='hide'}
state_summary_table <- asid %>%
  count(State, Incident.year, name = "yearly_count") %>% # count incidents per state per year
  group_by(State) %>%
  summarise(
    Total_Incidents = sum(yearly_count), # group by state to find max etc. 
    Peak_Year       = Incident.year[which.max(yearly_count)],
    Max_Incidents   = max(yearly_count),
    .groups = "drop"
  )

print(state_summary_table)

#NSW peaked in 2015 with 21 total incidents, 469 incidents total 
#QLD peaked in 1929 with 13 total incidents, 378 incidents total
#WA peaked in 2023 with 15 total incidents, 242 incidents total 
```

### Summary: Injury Type By State 

```{r,results='hide'}
state_injury<- asid %>%
  group_by(State, Victim.injury) %>%
  summarise(Total_Incidents = n(), .groups = "drop") %>% 
  arrange(State, desc(Total_Incidents))

print(state_injury)
```

### Plot One: Total Incidents Over Time (1791 - 2026); Faceted by State (NSW, QLD, and WA) 

```{r}
total_incident_plot <-asid %>%
  filter(State %in% c("NSW", "QLD", "WA")) %>% # filter to only include NSW, QLD, and WA
  ggplot(aes(x = Incident.year)) +
  geom_bar(fill = "steelblue", color = "steelblue") + 
  facet_wrap(~ State, ncol = 2) + 
  theme_minimal() +
  labs(
    title = "Raw Incident Counts Over Time by State",
    x = "Year",
    y = "Number of Incidents"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14), 
    strip.text = element_text(face = "bold", size = 11)
  )

print(total_incident_plot)

#ggsave(
 # filename = "Total_Incidents.png", 
  #plot = total_incident_plot,          
 # width = 10,               
 # height = 6,              
 # dpi = 300                
#)
```

### Plot Two: Total Incident Rates Per Capita; Faceted by State

```{r warning=FALSE}
# SUMMARISE: ASID data frame to only include incidents after 1900, in NSW, QLD, or WA only, and calculate total incidents per year 
# 1900 chosen as data very inconsistent earlier than 1900 (both in ASID and population data), plus public swimming laws at beaches relaxed in the early 1900's 

total_incidents_yearly <- asid %>%
  filter(
    Incident.year >= 1900, # any incidents that occurred in 1900 or later 
    State %in% c("NSW", "QLD", "WA") # top three states
  ) %>%
  group_by(State, Incident.year) %>%   # grouping by State and Year
  summarize(Total_Count = sum(n()), .groups = "drop") %>%
  complete(
    State, 
    Incident.year = 1900:2025, 
    fill = list(Total_Count = 0) #if any years where no incidents occur, fill with 0 (zeros are important!!)
  )

# Merge with population data and calculate the rate per 1 million people 
total_analysis_df <- total_incidents_yearly %>%
  left_join(final_population_df, by = c("Incident.year" = "Year", "State" = "State")) %>% # use left join to join by year and state
  mutate(Total_Rate_Per_Million = (Total_Count / Population) * 1000000) # calculate incident rate per one million people 

incident_rate_total_plot <-ggplot(total_analysis_df, aes(x = Incident.year, y = Total_Rate_Per_Million)) +
  geom_line(color = "grey50", linewidth = 0.5, alpha = 0.5) +   # raw total annual rates (semi-transparent grey line)
  geom_smooth(
    method = "glm", 
    method.args = list(family = "quasipoisson"), 
    formula = y ~ ns(x, df = 4), 
    se = FALSE, 
    color = "steelblue",  # aprroximated smoother trend line
    linewidth = 1.4
  ) +
  scale_x_continuous(limits = c(1900, 2026), breaks = seq(1900, 2026, by = 20)) + # fixed x-axis across all plots 
  facet_wrap(~ State, ncol = 2) + # fixed y-axis to not over inflate trends visually (all plots on the same y-axis)
  theme_minimal() +
  labs(
    title = "Total Shark Incident Rates Per Capita (1900 - 2025)",
    x = "Year",
    y = "Total Incident Rate (per 1 Million People)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )

print(incident_rate_total_plot)

#ggsave(
 # filename = "Total_Incidents_Rate.png", 
 # plot = incident_rate_total_plot,          
 # width = 10,               
 # height = 6,              
 # dpi = 300                
#)

```

### Create Milestones Dataframe with Shark Management Plans for each State 

```{r}
milestones <- data.frame(
  State = c("NSW", "QLD", "WA", "NSW"),
  Policy_Year = c(1937, 1962, 2012,2015),
  Label = c("BPP", "SCP", "SHMS","SMP") # Acronyms - Bather Protection Program (NSW), Shark Control Program (QLD), Shark Hazard Mitigation Strategy (WA), Shark Management Plan (NSW)
)
```

### Plot Three: Incident Outcomes Trends; Independent State Plots 

```{r}
incidents_yearly <- asid %>% 
  filter(
    Incident.year >= 1900,
    State %in% c("NSW", "QLD", "WA")
  ) %>%
  count(State, Incident.year, Victim.injury, name = "Count") %>% 
  complete(
    State, 
    Incident.year = 1900:2025,
    Victim.injury, 
    fill = list(Count = 0) # fills in zeros if years where certain outcomes did not occur. 
  )

analysis_df <- incidents_yearly %>%
  left_join(final_population_df, by = c("Incident.year" = "Year", "State" = "State")) %>%
  mutate(Rate_Per_Million = (Count / Population) * 1000000)

# If want to add raw data, add in the following code line 
#  geom_line(linewidth = 0.6, alpha = 0.4) +

plot_state_trends <- function(target_state, data = analysis_df, mstones = milestones) { # create custom function 
  state_data <- data %>% filter(State == target_state)   # Filter the main data set for just the chosen state
  state_milestones <- mstones %>% filter(State == target_state)   # Filter the milestones for just the chosen state
  
  ggplot(state_data, aes(x = Incident.year, y = Rate_Per_Million, color = Victim.injury)) +   # build the independent plot
    geom_smooth( # smoothed trend lines for each outcome type 
      aes(fill = Victim.injury), 
      method = "glm", 
      method.args = list(family = "quasipoisson"), 
      formula = y ~ ns(x, df = 4), 
      se = TRUE, 
      level = 0.95, # standard 95% confidence interval (buffer around each smoothed trend line)
      linewidth = 1.4,
      alpha = 0.2
    ) +
    geom_vline( # add vertical milestone lines at years when shark management began in each state 
      data = state_milestones, 
      aes(xintercept = Policy_Year), 
      linetype = "dashed", 
      color = "black", 
      alpha = 0.8,
      inherit.aes = FALSE
    ) +
    geom_text( # add milestone labels
      data = state_milestones, 
      aes(
        x = Policy_Year + 1.2, 
        y = 4, 
        label = Label
      ), 
      color = "black", 
      angle = 90, 
      vjust = 1.3, 
      hjust = 0.8,
      fontface = "bold.italic",
      inherit.aes = FALSE, 
      size = 3.5
    ) +
    scale_x_continuous(limits = c(1900, 2025), breaks = seq(1900, 2025, by = 10)) + # fixed x-axis and breaks
    scale_y_continuous(limits = c(0, 5), breaks = seq(0, 5, by = 1)) + # y-axis limit set to 5 for all states 
    theme_minimal() +
    scale_color_manual(
      values = c(
        "fatal" = "coral2",       
        "injured" = "dodgerblue3",   
        "uninjured" = "turquoise3"
      ),
      labels = c("fatal" = "Fatal", "injured" = "Injured", "uninjured" = "Uninjured")
    ) +
    scale_fill_manual( # for 95% CI buffer 
      values = c(
        "fatal" = "coral2",       
        "injured" = "dodgerblue3",   
        "uninjured" = "turquoise3"
      ),
      guide = "none" # hides duplicate legend
    ) +
    labs(
      title = paste(target_state, "- Population-Adjusted Shark Incident Rates (1900 - 2025)"),
      x = "Year",
      y = "Incident Rate (Per One Million People)",
      color = "Outcome"
    ) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

# Generate the NSW Plot
plot_nsw <- plot_state_trends("NSW")
print(plot_nsw)

# Generate the QLD Plot
plot_qld <- plot_state_trends("QLD")
print(plot_qld)

# Generate the WA Plot
plot_wa <- plot_state_trends("WA")
print(plot_wa)

ggsave(
  filename = "WA_injury.png", 
  plot = plot_wa,          
  width = 10,               
  height = 6,              
  dpi = 300                
)

ggsave(
  filename = "NSW_injury.png", 
  plot = plot_nsw,          
  width = 10,               
  height = 6,              
  dpi = 300                
)

ggsave(
  filename = "QLD_injury.png", 
  plot = plot_qld,          
  width = 10,               
  height = 6,              
  dpi = 300                
)
```

### Generate Era's Column in analysis_df

```{r}
analysis_df <- analysis_df %>%
  mutate(Era = case_when(
    State == "NSW" & Incident.year >= 1937 ~ "Post-Control", # NSW earliest shark management 
    State == "QLD" & Incident.year >= 1962 ~ "Post-Control",
    State == "WA"  & Incident.year >= 2012 ~ "Post-Control",
    TRUE ~ "Pre-Control"
  ),
  Era = factor(Era, levels = c("Pre-Control", "Post-Control")) # set levels so pre-control as baseline, rates will compare post-control to pre-control rates
  )
```

### GLM: Trends over Time 

```{r,results='hide'}
glm_trend_nsw <- glm(
  Count ~ Incident.year + offset(log(Population)),
  family = quasipoisson(),
  data = analysis_df %>% filter(State == "NSW")
)

glm_trend_qld <- glm(
  Count ~ Incident.year + offset(log(Population)),
  family = quasipoisson(),
  data = analysis_df %>% filter(State == "QLD")
)

glm_trend_wa <- glm(
  Count ~ Incident.year + offset(log(Population)),
  family = quasipoisson(),
  data = analysis_df %>% filter(State == "WA")
)

summary(glm_trend_nsw)
summary(glm_trend_qld)
summary(glm_trend_wa)
```

### New South Wales Statistical Analyses 

```{r, results='hide'}
## GLM for changes in incident outcomes 

analysis_df_nsw <- analysis_df %>%
  filter(State == "NSW") %>%
  mutate(
    Era = if_else(Incident.year >= 1937, "Post-Control", "Pre-Control"), # set NSW specific dates 
    Era = factor(Era, levels = c("Pre-Control", "Post-Control"))
  )

glm_fatal_nsw <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_nsw, Victim.injury == "fatal")
)

glm_injured_nsw <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_nsw, Victim.injury == "injured")
)

glm_uninjured_nsw <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_nsw, Victim.injury == "uninjured")
)

summary(glm_fatal_nsw)
summary(glm_injured_nsw)
summary(glm_uninjured_nsw)

# create custom function to reformat Incident Rate Ratio (IRR) and calculate percent changes
get_irr_table <- function(model, severity_name) {
  tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      Severity = severity_name,
      Pct_Change = (estimate - 1) * 100
    ) %>%
    select(Severity, Term = term, IRR = estimate, Pct_Change, p.value, conf.low, conf.high)
}

separate_models_summary_nsw <- bind_rows(
  get_irr_table(glm_fatal_nsw, "Fatalities"),
  get_irr_table(glm_injured_nsw, "Injured"),
  get_irr_table(glm_uninjured_nsw, "Uninjured")
)

print(separate_models_summary_nsw)
```

### Queensland Statistical Analyses 

```{r,results='hide'}
## GLM for changes in incident outcomes 

analysis_df_qld <- analysis_df %>%
  filter(State == "QLD") %>%
  mutate(
    # Set QLD-specific Era split
    Era = if_else(Incident.year >= 1962, "Post-Control", "Pre-Control"),
    # Ensure Pre-Control is the reference category
    Era = factor(Era, levels = c("Pre-Control", "Post-Control"))
  )

# fatal incidents model
glm_fatal_qld <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_qld, Victim.injury == "fatal")
)

# injured incidents model
glm_injured_qld <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_qld, Victim.injury == "injured")
)

# uninjured incidents model
glm_uninjured_qld <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_qld, Victim.injury == "uninjured")
)

summary(glm_fatal_qld)
summary(glm_injured_qld)
summary(glm_uninjured_qld)

# create custom function to reformat Incident Rate Ratio (IRR) and calculate percent changes
get_irr_table <- function(model, severity_name) {
  tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      Severity = severity_name,
      Pct_Change = (estimate - 1) * 100
    ) %>%
    select(Severity, Term = term, IRR = estimate, Pct_Change, p.value, conf.low, conf.high)
}

separate_models_summary_qld <- bind_rows(
  get_irr_table(glm_fatal_qld, "Fatalities"),
  get_irr_table(glm_injured_qld, "Injured"),
  get_irr_table(glm_uninjured_qld, "Uninjured")
)

print(separate_models_summary_qld)
```

### West Australia Statistical Analyses 

```{r,results='hide'}
## GLM for changes in incident outcomes 

analysis_df_wa<- analysis_df %>%
  filter(State == "WA") %>% #filter to only include WA incidents
  mutate(
    Era = if_else(Incident.year >= 2012, "Post-Control", "Pre-Control"), #set 2012 as the split between pre and post control
    Era = factor(Era, levels = c("Pre-Control", "Post-Control")) #ensure results are comparing post-control rates to pre-control rates
  )

# fatal incidents model 
glm_fatal_wa <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_wa, Victim.injury == "fatal")
)

# injured incidents model
glm_injured_wa <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_wa, Victim.injury == "injured")
)

# uninjured incidents model
glm_uninjured_wa <- glm(
  Count ~ Era + offset(log(Population)),
  family = quasipoisson(),
  data = filter(analysis_df_wa, Victim.injury == "uninjured")
)

summary(glm_fatal_wa)
summary(glm_injured_wa)
summary(glm_uninjured_wa)

# create custom function to reformat Incident Rate Ratio (IRR) and calculate percent changes
get_irr_table <- function(model, severity_name) {
  tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      Severity = severity_name,
      Pct_Change = (estimate - 1) * 100
    ) %>%
    select(Severity, Term = term, IRR = estimate, Pct_Change, p.value, conf.low, conf.high)
}

separate_models_summary_wa <- bind_rows(
  get_irr_table(glm_fatal_wa, "Fatalities"),
  get_irr_table(glm_injured_wa, "Injured"),
  get_irr_table(glm_uninjured_wa, "Uninjured")
)

print(separate_models_summary_wa)
```
