---
title: "Final Project"
author: "Caitlin Buckle"
date: "2026-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages 
```{r message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(splines)
```

Data downloaded from Taronga Conservation Society, Public Access Australian Shark Incident Database (ASID)
Accessed and downloaded, July 16th 2026 

Read ASID.csv into R
```{r}
asid<-read.csv("Data/ASID.csv")
summary(asid)
```

### Cleaning 

```{r}
asid%>% 
  count(State, sort = TRUE) #7 states/territories in dataset, three entries with no state/territory (need to remove), and one Qld (instead of QLD - will amend) 

asid$State[asid$State == "Qld"] <- "QLD" #changes Qld into QLD

asid<-asid%>%
  filter(!is.na(State) & State != "" & State != " ") #filters State column for any NA value, "" , or " " and removes

asid%>% 
  count(State, sort = TRUE) #State data is now all clean, with 7 states/territories with incidences in the database 
```

```{r}
asid%>% 
  count(Victim.injury, sort = TRUE) #'injured', 'fatal', and 'uninjured' the key categories, three for "blank" and two 'unknown' (will remove), one is 'injured' and one 'injury' (will include into 'injured')

asid$Victim.injury[asid$Victim.injury == "Injured"] <- "injured"
asid$Victim.injury[asid$Victim.injury == "injury"] <- "injured"

asid<-asid%>%
  filter(!is.na(Victim.injury) & Victim.injury != "" & Victim.injury != " " & Victim.injury != "unknown") #remove blank values and unknowns 

asid%>%
  count(Victim.injury,sort = TRUE) #now three categories - injured, fatal, or uninjured 
```

```{r}
summary(asid$Incident.year) #dataset starts in 1791 and latest data value entry was in 2026 

sum(is.na(asid$Incident.year)) #no incidences are missing year - great!! 

```

### Summary Tables - Total Incidents By State

```{r}
state_summary_table <- asid %>%
  count(State, Incident.year, name = "yearly_count") %>% #count incidences per state per year
  group_by(State) %>%
  summarise(
    Total_Incidents = sum(yearly_count), #group by state to find min, max etc. 
    Peak_Year       = Incident.year[which.max(yearly_count)],
    Max_Incidents   = max(yearly_count),
    .groups = "drop"
  )

print(state_summary_table)

#NSW peaked 2015 with 21 total incidences 
#QLD peaked 1929 with 13 total incidences 
#WA peaked 2023 with 15 total incidences
```

### Initial Plots 

```{r}
ggplot(asid, aes(x = Incident.year)) +
  geom_bar(fill = "steelblue", color = "steelblue") + 
  facet_wrap(~ State, ncol = 3) + 
  theme_minimal() +
  labs(
    title = "Incident Counts Over Time by State",
    x = "Year",
    y = "Number of Incidents"
  )

#can see that most incidents have occured in NSW, QLD, and WA 
#QLD had peak between 1925-1930 
#NSW peak recently ~2015
#WA also peaked recently 2023
```

```{r}
ggplot(asid, aes(x = Incident.year)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") + 
  facet_wrap(~ State, ncol = 3) + 
  theme_minimal() +
  labs(
    title = "Incident Trends Over Time by State (10-Year Bins)",
    x = "Year",
    y = "Number of Incidents"
  )

#grouped into ten-year bins so can more clearly visualise overall trends 
```

### Filter by Injury Type 

```{r}
state_injury<- asid %>%
  group_by(State, Victim.injury) %>%
  summarise(Total_Incidents = n(), .groups = "drop") %>%
  arrange(State, desc(Total_Incidents))

print(state_injury)
```

### Plot by State, with Separate Lines for Each Injury Type 

```{r}

asid %>%
  count(State, Incident.year, Victim.injury, name = "Incident_Count") %>%  #group and count incidents by state, year, and injury type
  ggplot(aes(x = Incident.year, y = Incident_Count, color = Victim.injury)) + 
  geom_line(linewidth = 0.8, alpha = 0.8) +
  facet_wrap(~ State, scales = "free_y", ncol = 3) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") + #high-contrast color palette
  labs(
    title = "Incident Injury Trends Over Time by State",
    x = "Year of Incident",
    y = "Number of Incidents",
    color = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11)
  )

## THIS IS SOOOOO HARD TO READ!! LOTS OF SPIKEY LINES AND OVERLAPS WHICH MAKES IT VERY DIFFICULT TO OBSERVE THE OVERALL TRENDS 
```

```{r warning=FALSE}
#Trialing smoothed lines 

asid %>%
  count(State, Incident.year, Victim.injury, name = "Incident_Count") %>%
  ggplot(aes(x = Incident.year, y = Incident_Count, color = Victim.injury)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.5, linewidth = 1) +
  coord_cartesian(ylim = c(0, NA)) + #to ensure that smoothed lines do not go further down than zero
  facet_wrap(~ State, scales = "free_y", ncol = 3) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Smoothed Incident Injury Trends by State (1791 - 2026)",
    subtitle = "Lines show the overall trend, filtering out year-to-year volatility",
    x = "Year",
    y = "Trend Value (Smoothed Counts)",
    color = "Injury Type"
  ) +
  theme(legend.position = "bottom")

#Definitely better, more readable however some of the trend lines dip below zero which is stupid 
```

```{r}
#ten-year bins approach 

ggplot(asid, aes(x = Incident.year, fill = Victim.injury)) +
  geom_histogram(binwidth = 10, color = "white", position = "stack") +
  facet_wrap(~ State, scales = "free_y", ncol = 3) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Total Incidents by State and Injury Type",
    subtitle = "Data grouped into 10-year time blocks",
    x = "Year",
    y = "Total Incidents in Block",
    fill = "Injury Type"
  ) +
  theme(legend.position = "bottom")

#getting better again, can see the overall trends in total incidences whilst also the changes injury types but still not quite right 
```

```{r}
# NSW only for all years (1791 - 2026)

nsw_data <- asid %>% 
  filter(State == "NSW") #dataframe with NSW incidences only 

ggplot(nsw_data, aes(x = Incident.year, fill = Victim.injury)) + #bars of count per incident year, filled by injury type count 
  geom_histogram(
    binwidth = 5, 
    color = "white",      
    linewidth = 0.2,       
    position = "stack"  #stacked by injury type count   
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "NSW Shark Incidences by Injury Type (1791 - 2026)",
    subtitle = "Data grouped into 5-year intervals",
    x = "Year",
    y = "Total Incidents per 5-Year Period",
    fill = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

```{r}
# QLD only for all years (1791 - 2026)

qld_data <- asid %>% 
  filter(State == "QLD") #dataframe with NSW incidences only 

ggplot(qld_data, aes(x = Incident.year, fill = Victim.injury)) + #bars of count per incident year, filled by injury type count 
  geom_histogram(
    binwidth = 5, 
    color = "white",      
    linewidth = 0.2,       
    position = "stack"  #stacked by injury type count   
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "QLD Shark Incidences by Injury Type (1791 - 2026)",
    subtitle = "Data grouped into 5-year intervals",
    x = "Year",
    y = "Total Incidents per 5-Year Period",
    fill = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

```{r}
#WA only for all years (1791 - 2026)

wa_data <- asid %>% 
  filter(State == "WA") #dataframe with NSW incidences only 

ggplot(wa_data, aes(x = Incident.year, fill = Victim.injury)) + #bars of count per incident year, filled by injury type count 
  geom_histogram(
    binwidth = 5, 
    color = "white",      
    linewidth = 0.2,       
    position = "stack"  #stacked by injury type count   
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "WA Shark Incidences by Injury Type (1791 - 2026)",
    subtitle = "Data grouped into 5-year intervals",
    x = "Year",
    y = "Total Incidents per 5-Year Period",
    fill = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )

```


### Refining Plots 

```{r}
#NSW - after 1850 and highlighted when active shark management started 

nsw_filtered <- asid %>%
  filter(
    State == "NSW",
    Incident.year >= 1850
  ) %>%
  count(Incident.year, name = "Total_Incidents") %>%
  mutate(Era = if_else(Incident.year > 1937, "After 1937", "1850 - 1937"))


ggplot(nsw_filtered, aes(x = Incident.year, y = Total_Incidents, fill = Era)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("1851 - 1937" = "#A0A0A0", "After 1937" = "#2B6CB0")) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1850, 2026, by = 20)) +
  labs(
    title = "NSW Yearly Incident Counts (1851 - 2026)",
    subtitle = "Highlighting the presence of active shark management plans",
    x = "Year",
    y = "Number of Incidents",
    fill = "Time Period"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

```{r}
#QLD - filtered to incidences after 1850, and 1962 onwards highlighted as years with active management 

qld_filtered <- asid %>%
  filter(
    State == "QLD",
    Incident.year >= 1850
  ) %>%
  count(Incident.year, name = "Total_Incidents") %>%
  mutate(Era = if_else(Incident.year > 1962, "After 1962", "1850 - 1962"))


ggplot(qld_filtered, aes(x = Incident.year, y = Total_Incidents, fill = Era)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("1850 - 1962" = "#A0A0A0", "After 1962" = "#2B6CB0")) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1850, 2026, by = 20)) +
  labs(
    title = "QLD Yearly Incident Counts (1850 - 2026)",
    subtitle = "Highlighting the presence of active shark management plans",
    x = "Year",
    y = "Number of Incidents",
    fill = "Time Period"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

```{r}
#WA - filtered to incidences after 1850, and 2012 onwards highlighted as years with active management 

qld_filtered <- asid %>%
  filter(
    State == "WA",
    Incident.year >= 1850
  ) %>%
  count(Incident.year, name = "Total_Incidents") %>%
  mutate(Era = if_else(Incident.year > 2012, "After 2012", "1850 - 2012"))


ggplot(qld_filtered, aes(x = Incident.year, y = Total_Incidents, fill = Era)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("1850 - 2012" = "#A0A0A0", "After 2012" = "#2B6CB0")) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1850, 2026, by = 20)) +
  labs(
    title = "WA Yearly Incident Counts (1850 - 2026)",
    subtitle = "Highlighting the presence of active shark management plans",
    x = "Year",
    y = "Number of Incidents",
    fill = "Time Period"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

### TO DO 

Filted ASID state-specific dataframes to 1900 onwards 

Omit attacks from 2026 as the year is not over and we do not have specific population statistics 

Find census data from 1900 - present for each state from Australian Bureau of Statistics and read into R as a csv 

### Population Data 

1901 - 2011: ASB 3105.0.65.001 Australian Historical Population Statistics, 2014 Table 1.2 Population by sex, states, and territories, 1901 onwards
2012 - 2019: ASB 3101.0 National State and Territory Population, Table 4: Estimated Resident Population, States and Territories 
2020 - 2025: ABS 3101.001.202512 National, State, and Territory Population, Dec 2025  Table 5: Population by sex-States and Territories

```{r}
#Clean the 1901-2011 file, horizontal orientation (State on y, Year on x)
pop_1901_2011<- read_csv("Data/1901_2011.csv")

clean_1901_2011 <- pop_1901_2011 %>% 
  rename(State = Year) %>% #need to rename the Year column to State (incorrect heading from original .csv file)
    pivot_longer(   #pivot all the numeric year columns down, leaving State alone
    cols = -State,               
    names_to = "Year", 
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(Year))

pop_2012_2019 <- read_csv("Data/2012_2019.csv")
pop_2020_2025 <- read_csv("Data/2020_2025.csv")

clean_2012_2019 <- pop_2012_2019 %>%
  pivot_longer(
    cols = -Year,                
    names_to = "State", 
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(Year))

clean_2020_2025 <- pop_2020_2025 %>%
  pivot_longer(
    cols = -Year,                
    names_to = "State", 
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(Year))

final_population_df <- bind_rows( #join together the rows in the separate files into one data frame with all population data 1901-2025
    clean_1901_2011, 
    clean_2012_2019, 
    clean_2020_2025
  ) %>%
  mutate(State = toupper(State)) %>% #ensure column headers match the ASID headers to avoid confusion 
  arrange(State, Year)

summary(final_population_df)
```

### INCIDENTS PER CAPITA 

```{r}
#Convert shark attack incidents per year into shark incidents per year per capita to account for changes in population 

# =======================================================
# 1. Summarize incident data by State, Year, and Injury
# =======================================================
incidents_yearly <- asid %>%
  filter(
    Incident.year >= 1900,
    State %in% c("NSW", "QLD", "WA")
  ) %>%
  count(State, Incident.year, Victim.injury, name = "Count")

# =======================================================
# 2. Merge with your new master population dataframe
# =======================================================
analysis_df <- incidents_yearly %>%
  # Merging using the exact column names we structured earlier
  left_join(final_population_df, by = c("Incident.year" = "Year", "State" = "State")) %>%
  
  # Calculate the population-adjusted rate per 1 million people
  mutate(Rate_Per_Million = (Count / Population) * 1000000)

# =======================================================
# 3. Add the Active Management Flags (WA updated to 2012)
# =======================================================
analysis_df <- analysis_df %>%
  mutate(Management = case_when(
    State == "NSW" & Incident.year >= 1937 ~ "Active Management (Nets)",
    State == "QLD" & Incident.year >= 1962 ~ "Active Management (Nets/Drumlines)",
    State == "WA"  & Incident.year >= 2012 ~ "Modern Mitigation Era",
    TRUE ~ "No Active Program"
  ))

# =======================================================
# 4. Generate the Comparative Visualization
# =======================================================
# Define the intervention milestone vertical lines (WA matching at 2012)
milestones <- data.frame(
  State = c("NSW", "QLD", "WA"),
  Policy_Year = c(1937, 1962, 2012),
  Label = c("Nets Introduced", "Nets/Drumlines", "Mitigation Era")
)

ggplot(analysis_df, aes(x = Incident.year, y = Rate_Per_Million, color = Victim.injury)) +
  
  # Non-negative smoothing model to beautifully trend the per-capita rate
  geom_smooth(
    method = "glm", 
    method.args = list(family = "quasipoisson"), 
    formula = y ~ ns(x, df = 4), 
    se = FALSE, 
    linewidth = 1
  ) +
  
  # Add the vertical milestone lines
  geom_vline(
    data = milestones, 
    aes(xintercept = Policy_Year), 
    linetype = "dashed", 
    color = "black", 
    alpha = 0.6
  ) +
  
  # Label the lines clearly on the plots
  geom_text(
    data = milestones, 
    aes(x = Policy_Year + 2, y = Inf, label = Label), 
    color = "black", 
    angle = 90, 
    vjust = 1.5, 
    hjust = 1.1, 
    inherit.aes = FALSE, 
    size = 3
  ) +
  
  # Separate charts for each state to track regional differences
  facet_wrap(~ State, scales = "free_y", ncol = 3) +
  
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Population-Adjusted Shark Incident Rates over Time (1900 - 2026)",
    subtitle = "Rates reflect incidents per 1,000,000 people. Dashed lines indicate active management rollouts.",
    x = "Year",
    y = "Incident Rate (per 1 Million People)",
    color = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11)
  )
```

```{r}
#per capita graph w/ raw data 
library(dplyr)
library(ggplot2)

# =======================================================
# 1. Filter and summarize incident data
# =======================================================
incidents_yearly <- asid %>%
  filter(
    Incident.year >= 1900,
    State %in% c("NSW", "QLD", "WA")
  ) %>%
  count(State, Incident.year, Victim.injury, name = "Count")

# =======================================================
# 2. Merge with population dataframe & calculate rates
# =======================================================
analysis_df <- incidents_yearly %>%
  left_join(final_population_df, by = c("Incident.year" = "Year", "State" = "State")) %>%
  mutate(Rate_Per_Million = (Count / Population) * 1000000)

# =======================================================
# 3. Generate the Non-Smoothed (Raw Line) Visualization
# =======================================================
milestones <- data.frame(
  State = c("NSW", "QLD", "WA"),
  Policy_Year = c(1937, 1962, 2012),
  Label = c("Nets Introduced", "Nets/Drumlines", "Mitigation Era")
)

ggplot(analysis_df, aes(x = Incident.year, y = Rate_Per_Million, color = Victim.injury)) +
  
  # Simply draw raw lines connecting the exact data points for each year
  geom_line(linewidth = 0.8, alpha = 0.8) +
  
  # Add the vertical milestone lines
  geom_vline(
    data = milestones, 
    aes(xintercept = Policy_Year), 
    linetype = "dashed", 
    color = "black", 
    alpha = 0.6
  ) +
  
  # Label the lines clearly on the plots
  geom_text(
    data = milestones, 
    aes(x = Policy_Year + 2, y = Inf, label = Label), 
    color = "black", 
    angle = 90, 
    vjust = 1.5, 
    hjust = 1.1, 
    inherit.aes = FALSE, 
    size = 3
  ) +
  
  # Generates exactly 3 charts side-by-side
  facet_wrap(~ State, scales = "free_y", ncol = 3) +
  
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "RAW Population-Adjusted Shark Incident Rates (1900 - 2026)",
    subtitle = "Spikes represent high per-capita volatility in individual years",
    x = "Year",
    y = "Incident Rate (per 1 Million People)",
    color = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11)
  )
```

```{r}
#code with trendlines 

# =======================================================
# 1. Filter and summarize incident data
# =======================================================
incidents_yearly <- asid %>%
  filter(
    Incident.year >= 1900,
    State %in% c("NSW", "QLD", "WA")
  ) %>%
  count(State, Incident.year, Victim.injury, name = "Count")

# =======================================================
# 2. Merge with population dataframe & calculate rates
# =======================================================
analysis_df <- incidents_yearly %>%
  left_join(final_population_df, by = c("Incident.year" = "Year", "State" = "State")) %>%
  mutate(Rate_Per_Million = (Count / Population) * 1000000)

# =======================================================
# 3. Generate the Overlay Visualization
# =======================================================
milestones <- data.frame(
  State = c("NSW", "QLD", "WA"),
  Policy_Year = c(1937, 1962, 2012),
  Label = c("Nets Introduced", "Nets/Drumlines", "Mitigation Era")
)

ggplot(analysis_df, aes(x = Incident.year, y = Rate_Per_Million, color = Victim.injury)) +
  
  # LAYER 1: Raw yearly spikes (made semi-transparent with alpha = 0.4)
  geom_line(linewidth = 0.5, alpha = 0.4) +
  
  # LAYER 2: Strong, zero-constrained mathematical trendlines over the top
  geom_smooth(
    method = "glm", 
    method.args = list(family = "quasipoisson"), 
    formula = y ~ ns(x, df = 4), 
    se = FALSE, 
    linewidth = 1.3
  ) +
  
  # Add the vertical milestone lines
  geom_vline(
    data = milestones, 
    aes(xintercept = Policy_Year), 
    linetype = "dashed", 
    color = "black", 
    alpha = 0.6
  ) +
  
  # Label the lines clearly on the plots
  geom_text(
    data = milestones, 
    aes(x = Policy_Year + 2, y = Inf, label = Label), 
    color = "black", 
    angle = 90, 
    vjust = 1.5, 
    hjust = 1.1, 
    inherit.aes = FALSE, 
    size = 3
  ) +
  
  # Generates exactly 3 charts side-by-side
  facet_wrap(~ State, ncol = 3) +
  
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Population-Adjusted Shark Incident Rates with Long-Term Trends",
    subtitle = "Faded lines show raw annual rates; bold curves highlight non-negative smoothed trends.",
    x = "Year",
    y = "Incident Rate (per 1 Million People)",
    color = "Injury Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11)
  )
```

## THOUGHTS 

looks pretty good, but would it be better to have the plots independent so the y-axis are specific (could have them specific in faceted but then can mislead by making smaller trends e.g., NSW look big when its actually the smallest)

for the presentation - have them separated and some information below each plot with key info, max, min, management plan inception etc. general trend etc. 
